ProbabilityBoundsAnalysis.jl:\\ Arithmetic with sets of distributions

### 2022-04-14 paper/paper.tex
refactor introduction text

% JuliaCon proceedings template
\documentclass{juliacon}
\setcounter{page}{1}
\usepackage{caption}
\usepackage{amsmath}
\usepackage{xcolor}
%\bibliographystyle{juliacon}
\usepackage[labelformat=simple]{subfig}
\renewcommand\thesubfigure{\arabic{subfigure}}
\renewcommand{\subfigurename}{Figure}
\begin{document}
\input{header}
\maketitle
\begin{abstract}
Probability bounds analysis combines interval arithmetic with probability theory, and provides a representation of sets of distributions in structures called probability boxes (p-boxes). P-boxes generalise both distribution functions and intervals, and return interval bounds on all probabilistic quantities, for example sample realisations, cdfs, and probability measures are all intervals. This framework also allows for the comprehensive propagation of probabilities through calculations in a rigorous way, in a similar fashion that interval arithmetic does for sets of real values. \texttt{ProbabilityBoundsAnalysis.jl} provides a rigorous arithmetic of random variables, where both marginal (univariate) distributions and dependency information can be known, partially known or missing entirely. We describe the main theoretical elements of probability bounds analysis, and provide a simplified implementation of the method in code snippets which can be readily evaluated in the Julia command terminal.
\end{abstract}
\section{Introduction}
\label{sec:intro}
An arithmetic of probability distributions has long held interest among mathematicians and scientists. Indeed it was Kolmogorov who originally asked about the result of an operation between two random variables without knowing their joint distribution. This was answered for the sum by Makarov~\cite{makarov1982estimates}, who showed the result was a set of distributions and was able to provide bounds on this function. Sklar, Schweizer and Frank generalised this result to other positive binary operations~\cite{frank1987best,schweizer2011probabilistic}. In this pursuit they created copulas, a general way to encode probabilistic dependence independently from marginals. These are now an essential object used in probabilistic modelling. In his dissertation~\cite{williamson1989probabilistic}, Williamson described an algorithm for efficiently performing these arithmetic operations, which can give guaranteed bounds on probability distributions in terms of an upper and lower cumulative distribution function (cdf). He called his method \textit{probabilistic arithmetic} and his sets of distributions \textit{dependency bounds}. Since then the method has been generalised~\cite{ferson2015constructing,ferson1996whereof,ferson2004arithmetic} to most of the base binary and unary operations that would be present in a programming language. Probability boxes (p-boxes) are the name now given to these structures.

@@

The idea of bounding probability has a long tradition throughout the history of probability theory. George Boole~\cite{boole1854investigation, hailperin1986boole} used the notion of interval bounds on probability. Chebyshev~\cite{chebyshev1874valeurs} described bounds on a distribution when only the mean and variance of the variable are known, and Markov~\cite{markoff1900question} found bounds on a positive variable when only the mean is known.  Fréchet~\cite{frechet1935generalisation} discovered how to bound joint distributions solely from knowing the marginal distributions, without making independence assumptions. Bounding probabilities has continued to the present day, culminating into the modern theory of imprecise probabilities~\cite{walley1991statistical, klir2013uncertainty, troffaes2014lower, augustin2014introduction}.
\begin{figure*}[htp]

@@

\end{figure*}
Imprecise probability is effectively a generalisation of probability theory where uncertainty can be expressed about the probability measure. This is particularly relevant when information is scarce, unreliable, vague, conflicting or imprecise. In such cases defining a unique probability distribution is difficult. P-boxes are one of many ways to describe a set of distributions. Others include: Dempster-Shafer structures~\cite{dempster2008upper,shafer1976mathematical}, random sets~\cite{molchanov2005theory}, possibility distributions~\cite{zadeh1978fuzzy,dubois1988possibility, hose2019possibilistic}, lower previsions~\cite{troffaes2014lower}, and credal sets~\cite{levi1983enterprise}. These structures were discovered independently, but are often synonymous and can be translated from one to another, with different degrees of generality. Imprecise probability unifies the underlying theories. For a comprehensive overview of the theory, and a formal description of uncertainty and information in terms of these structures, \cite{klir2013uncertainty} is recommended.
\subsection{Automatic uncertainty in a programming language: why Julia?}
The long term goal of probability bounds analysis is to create a fully uncertain programming language, where any computer variable may be represented as an interval, distribution, p-box or other imprecise quantity. Such a framework would allow for uncertain extensions of deterministic functions to be computed in an automatic, rigorous, and tight fashion. Those with little knowledge of probability theory could add uncertainty to their computations with minimal effort. Stochastic computation often require specialist training to be used effectively, and are usually not rigorous and are famously computationally expensive. \texttt{ProbabilityBoundsAnalysis.jl} provides the arithmetic basis for such a framework. In this section we give some arguments why Julia~\cite{bezanson2017julia} is an ideal target language for such a framework.

@@

\subsubsection{Metaprogramming}
An uncertain programming language would require \textit{source-to-source} transformations at times. For example, one way to evaluate a programs control-flow (\texttt{if-else} conditions) with uncertain inputs is to use metaprogramming. The programs control flow could be unraveled, and with each possible conditional branch evaluated for the possible instance of the variable, and then combined as an output. For example consider the following function
\begin{lstlisting}[language = Julia]
  function unc_flow(x)
    if x < 0 
      return x - 1
    else 
      return x
    end
  end
\end{lstlisting}
\noindent An interval evaluation of $\text{unc\_flow}([-1,1])$ would be the union of the output of the different branches, the interval union $[-2, -1) \cup [0, 1]$. For a language like Python, this would require an uncertainty pre-compiler (like that proposed in~\cite{gray2019computing}) whereby the python code is first parsed using a separate program, and then evaluated in Python. In Julia, this can be handled natively with its metaprogramming capabilities.
\begin{figure*}[htp]

@@

A bivariate copula (2-copula) $C$ is any function $C:[0,1]^2 \rightarrow [0,  1]$ with the following properties
\begin{enumerate}
  \item Grounded: $C(0,v) = C(u,0) = 0$,
  \item Uniform margins: $C(u,1) = u; \;C(1,v) = v$,
  \item 2-increasing: \\ $C(u_{2},v_{2}) - C_{}(u_{2}, v_{1}) - C(u_{1}, v_{2}) + C(u_{1}, v_{1}) \ge 0$\\ for all $0 \le u_{1} \le u_{2} \le 1$ and $0 \le v_{1} \le v_{2} \le 1$ .
\end{enumerate}
\noindent Three important 2-copulas are
\begin{align*}
  W(u,v) &= \mathrm{max}( u + v-1,0),  \\
  \Pi(u,v) &= uv, \\
  M(u,v) &= \mathrm{min}(u,v),
\end{align*}
\noindent where $W$ encodes maximally negative correlation, $\Pi$ encodes stochastic independence, and $M$ encodes maximally positive correlation. Moreover, $W$ and $M$ are bounds on all 2-copulas: $W(u,v) \leq C(u,v) \leq M(u,v)$. A bivariate cdf $F_{XY}$ can be constructed from two marginal cdfs $F_{X}$ and $F_{Y}$ and copula $C$ using Sklar's theorem
\begin{equation*}
  F_{XY}(x,y) = C(F_{X}(x), F_{Y}(y)),
\end{equation*}
\noindent where the constructed bivariate distribution will follow the dependence encoded in the copula and have the prescribed marginals. Sklar's theorem also works in reverse, where the dependence can be extracted from a multivariate distribution: 
\begin{equation*}
  C(u,v) = F_{XY}(F^{-1}_{X}(u), F^{-1}_{Y}(v)).
\end{equation*}
\noindent Copulas also give a very convenient way to sample from $F_{XY}$ in a generalisation of inverse transform sampling, where first the copula is sampled $(u,v) \sim C$, and then transformed through the inverses of the marginals $(F^{-1}_{X}(u), F^{-1}_{Y}(v))$.
\begin{figure*}[t!]
  \centering
  \includegraphics[width=0.45\textwidth]{../examples/JuliaCon/fig7/biv_cdf.pdf}\hfill
  \includegraphics[width=0.45\textwidth]{../examples/JuliaCon/fig7/samples.pdf}
  \caption{Cdf bounds of a bivariate p-box (right) with $X \sim \text{Beta}([4,8], 3)$, $Y \sim \text{N}([0,2], 2)$, and $\rho_{XY} = -0.8$, and 100 samples (left).}
  \label{fig:figure7}
\end{figure*}

Sklar's theorem has a straightforward imprecise extension~\cite{montes2015sklar}, where the two bivariate bounds of a p-box $[\underline{F}_{XY}(x,y), \overline{F}_{XY}(x,y)]$ can expanded in terms of two marginal p-boxes and bounds on a copula $[\underline{C}(u,v), \overline{C}(u,v)]$
\begin{align*}
  \underline{F}_{XY}(x,y) &= \underline{C}(\underline{F}_{X}(x), \underline{F}_{Y}(y)), \\ 
  \overline{F}_{XY}(x,y) &= \overline{C}(\overline{F}_{X}(x), \overline{F}_{Y}(y)).
\end{align*}
Samples and the probability measure can also be found from a bivariate p-box. Like how a sample of a univariate p-box is an interval, a sample of a bivariate p-box will be an interval box, i.e. a two dimensional rectangular set. Multiple samples will therefore be a population of interval boxes, which together follow the dependency encoded in $C$. The bottom of Figure \ref{fig:figure7} shows 100 samples of a bivariate p-box, whose marginals are $X \sim \text{Beta}([4,8], 3)$ and $Y \sim \text{N}([0,2], 2)$, and with the Gaussian copula (multivariate normal's copula) with correlation coefficient $\rho_{XY} = -0.8$: $C = \Phi_{-0.8}$. The top of Figure \ref{fig:figure7} shows the cdf bounds. The probability measure from a bivariate p-box in the rectangular set $ U = [x_{1}, x_{2}] \times [y_{1}, y_{2}]$ is
\begin{align*}
  \underline{\mathbb{P}}(U) &= \text{max}(0, \underline{F}(x_{2}, y_{2}) - \overline{F}(x_{1}, y_{2}) - \overline{F}(x_{2}, y_{1}) + \underline{F}(x_{1}, y_{1})) ,\\ 
  \overline{\mathbb{P}}(U)  &= \text{min}(1, \overline{F}(x_{2}, y_{2}) - \underline{F}(x_{1}, y_{2}) - \underline{F}(x_{2}, y_{1}) + \overline{F}(x_{1}, y_{1}) ).
\end{align*}
\begin{figure*}[htp]

>>>>

% JuliaCon proceedings template
\documentclass{juliacon}
\setcounter{page}{1}
\usepackage{caption}
\usepackage{amsmath}
\usepackage{xcolor}
%\bibliographystyle{juliacon}
\usepackage[labelformat=simple]{subfig}
\renewcommand\thesubfigure{\arabic{subfigure}}
\renewcommand{\subfigurename}{Figure}
\begin{document}
\input{header}
\maketitle
\begin{abstract}
Probability bounds analysis combines interval arithmetic with probability theory, and provides a representation of sets of distributions in structures called probability boxes (p-boxes). P-boxes generalise both distribution functions and intervals, and return interval bounds on all probabilistic quantities, for example sample realisations, cdfs, and probability measures are all intervals. This framework also allows for the comprehensive propagation of probabilities through calculations in a rigorous way, in a similar fashion that interval arithmetic does for sets of real values. \texttt{ProbabilityBoundsAnalysis.jl} provides a rigorous arithmetic of random variables, where both marginal (univariate) distributions and dependency information can be known, partially known or missing entirely. We describe the main theoretical elements of probability bounds analysis, and provide a simplified implementation of the method in code snippets which can be readily evaluated in the Julia command terminal.
\end{abstract}
\section{Introduction}
\label{sec:intro}
An arithmetic of probability distributions has long held interest among mathematicians and scientists. Indeed it was Kolmogorov who originally asked about the result of an operation between two random variables without knowing their joint distribution. This was answered for the sum by Makarov~\cite{makarov1982estimates}, who showed the result was a set of distributions and was able to provide bounds on this function. Sklar, Schweizer and Frank generalised this result to other positive binary operations~\cite{frank1987best,schweizer2011probabilistic}. In this pursuit they created copulas, a general way to encode probabilistic dependence independently from marginals, which are now essential objects in probabilistic modelling. Williamson and Downs~\cite{williamson1990probabilistic} described an algorithm for efficiently performing these arithmetic operations, which can give guaranteed bounds on probability distributions in terms of an upper and lower cumulative distribution function (cdf). They called their method \textit{probabilistic arithmetic} and their sets of distributions \textit{dependency bounds}. Since then the method has been generalised~\cite{ferson2015constructing,ferson1996whereof,ferson2004arithmetic} to most of the base binary and unary operations that would be present in a programming language. Probability boxes (p-boxes) are the name now given to these structures.

@@

The idea of bounding probability has a long tradition throughout the history of probability theory. George Boole~\cite{boole1854investigation, hailperin1986boole} used the notion of interval bounds on probability. Chebyshev~\cite{chebyshev1874valeurs} described bounds on a distribution when only the mean and variance of the variable are known, and Markov~\cite{markoff1900question} found bounds on a positive variable when only the mean is known. Fréchet~\cite{frechet1935generalisation} discovered how to bound joint distributions solely from knowing the marginal distributions, without making (in)dependence assumptions. Bounding probabilities has continued to the present day, culminating into the modern theory of imprecise probabilities~\cite{walley1991statistical, klir2013uncertainty, troffaes2014lower, augustin2014introduction}.
\begin{figure*}[htp]

@@

\end{figure*}
Imprecise probability is effectively a generalisation of probability theory where uncertainty can be expressed about the probability measure. This is particularly relevant when information is scarce, unreliable, vague, conflicting, or imprecise: cases where defining a unique probability distribution is difficult. P-boxes are one of many ways to describe a set of distributions. Others include: Dempster-Shafer structures~\cite{dempster2008upper,shafer1976mathematical}, random sets~\cite{molchanov2005theory}, possibility distributions~\cite{zadeh1978fuzzy,dubois1988possibility, hose2019possibilistic}, lower previsions~\cite{troffaes2014lower}, and credal sets~\cite{levi1983enterprise}. These structures were discovered independently, but are often synonymous (with different degrees of generality) and can be transformed from one to another. Imprecise probability unifies the underlying theories. For an overview of the theory, and a formal description of uncertainty and information in terms of these structures, work by George Klir~\cite{klir2013uncertainty} is recommended.
\subsection{Automatic uncertainty in a programming language: why Julia?}
A potential long term goal of probability bounds analysis is to create a uncertain programming language, where any computer variable may be represented as an interval, distribution, p-box, or other imprecise quantity. Such a framework would allow for uncertain extensions of deterministic functions to be computed in an automatic, rigorous, and tight fashion. Those with little knowledge of probability theory could add uncertainty to their computations with minimal effort. Stochastic computation often require specialist training to be used effectively, and are usually not rigorous and are famously computationally expensive. \texttt{ProbabilityBoundsAnalysis.jl} provides the arithmetic basis for such a framework. In this section we give some arguments why Julia~\cite{bezanson2017julia} is an ideal target language for such a framework.

@@

\subsubsection{Metaprogramming}
An uncertain programming language would require \textit{source-to-source} transformations at times. For example, one way to evaluate a programs control-flow (\texttt{if-else} conditions) with uncertain inputs is to use metaprogramming. The programs control flow could be unravelled, and with each possible conditional branch evaluated for the possible instance of the variable, and then combined as an output. For example consider the following function
\begin{lstlisting}[language = Julia]
  function unc_flow(x)
    if x < 0 
      return x - 1
    else 
      return x
    end
  end
\end{lstlisting}
\noindent An interval evaluation of $\text{unc\_flow}([-1,1])$ would be the union of the output of the different branches, the interval union $[-2, -1) \cup [0, 1]$. For a language like Python, this would require an uncertainty pre-compiler (like that proposed in~\cite{gray2019computing}) whereby the python code is first parsed using a separate program, and then evaluated in Python. In Julia, this can be handled natively with its metaprogramming capabilities.
\begin{figure*}[htp]

@@

A bivariate copula (2-copula) $C$ is any function $C:[0,1]^2 \rightarrow [0,  1]$ with the following properties
\begin{enumerate}
  \item Grounded: $C(0,v) = C(u,0) = 0$,
  \item Uniform margins: $C(u,1) = u; \;C(1,v) = v$,
  \item 2-increasing: \\ $C(u_{2},v_{2}) - C_{}(u_{2}, v_{1}) - C(u_{1}, v_{2}) + C(u_{1}, v_{1}) \ge 0$\\ for all $0 \le u_{1} \le u_{2} \le 1$ and $0 \le v_{1} \le v_{2} \le 1$ .
\end{enumerate}
\noindent Three important 2-copulas are
\begin{align*}
  W(u,v) &= \mathrm{max}( u + v-1,0),  \\
  \Pi(u,v) &= uv, \\
  M(u,v) &= \mathrm{min}(u,v),
\end{align*}
\noindent where $W$ encodes maximally negative correlation, $\Pi$ encodes stochastic independence, and $M$ encodes maximally positive correlation. Moreover, $W$ and $M$ are bounds on all 2-copulas: $W(u,v) \leq C(u,v) \leq M(u,v)$. A bivariate cdf $F_{XY}$ can be constructed from two marginal cdfs $F_{X}$ and $F_{Y}$ and copula $C$ using Sklar's theorem
\begin{equation*}
  F_{XY}(x,y) = C(F_{X}(x), F_{Y}(y)),
\end{equation*}
\noindent where the constructed bivariate distribution will follow the dependence encoded in the copula and have the prescribed marginals. Sklar's theorem also works in reverse, where the dependence can be extracted from a multivariate distribution: 
\begin{equation*}
  C(u,v) = F_{XY}(F^{-1}_{X}(u), F^{-1}_{Y}(v)).
\end{equation*}
\noindent Copulas also give a very convenient way to sample from $F_{XY}$ in a generalisation of inverse transform sampling, where first samples are drawn from the copula $(u,v) \sim C$, and then transformed through the inverses of the marginals $(F^{-1}_{X}(u), F^{-1}_{Y}(v))$.
\begin{figure*}[t!]
  \centering
  \includegraphics[width=0.45\textwidth]{../examples/JuliaCon/fig7/biv_cdf.pdf}\hfill
  \includegraphics[width=0.45\textwidth]{../examples/JuliaCon/fig7/samples.pdf}
  \caption{Cdf bounds of a bivariate p-box (right) with $X \sim \text{Beta}([4,8], 3)$, $Y \sim \text{N}([0,2], 2)$, and $\rho_{XY} = -0.8$, and 100 samples (left).}
  \label{fig:figure7}
\end{figure*}

Sklar's theorem has a imprecise extension~\cite{montes2015sklar}, where the two bivariate bounds of a p-box $[\underline{F}_{XY}(x,y), \overline{F}_{XY}(x,y)]$ can expanded in terms of two marginal p-boxes and bounds a copula $[\underline{C}(u,v), \overline{C}(u,v)]$
\begin{align*}
  \underline{F}_{XY}(x,y) &= \underline{C}(\underline{F}_{X}(x), \underline{F}_{Y}(y)), \\ 
  \overline{F}_{XY}(x,y) &= \overline{C}(\overline{F}_{X}(x), \overline{F}_{Y}(y)).
\end{align*}
Samples and the probability measure can also be found from a bivariate p-box. Like how a sample of a univariate p-box is an interval, a sample of a bivariate p-box will be an interval box, i.e. a two dimensional rectangular set. Multiple samples will therefore be a population of interval boxes, which together follow the dependency encoded in $C$. The bottom of Figure \ref{fig:figure7} shows 100 samples of a bivariate p-box, whose marginals are $X \sim \text{Beta}([4,8], 3)$ and $Y \sim \text{N}([0,2], 2)$, and with the Gaussian copula (multivariate normal's copula) with correlation coefficient $\rho_{XY} = -0.8$: $C = \Phi_{-0.8}$. The top of Figure \ref{fig:figure7} shows the cdf bounds. The probability measure from a bivariate p-box in the rectangular set $ U = [x_{1}, x_{2}] \times [y_{1}, y_{2}]$ is
\begin{align*}
  \underline{\mathbb{P}}(U) &= \text{max}(0, \underline{F}(x_{2}, y_{2}) - \overline{F}(x_{1}, y_{2}) - \overline{F}(x_{2}, y_{1}) + \underline{F}(x_{1}, y_{1})) ,\\ 
  \overline{\mathbb{P}}(U)  &= \text{min}(1, \overline{F}(x_{2}, y_{2}) - \underline{F}(x_{1}, y_{2}) - \underline{F}(x_{2}, y_{1}) + \overline{F}(x_{1}, y_{1}) ).
\end{align*}
\begin{figure*}[htp]
